---
title: "North and Central American Mammals Functional Rarity Read Me"
author: "Matthias Grenié"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: rmarkdown::github_document
always_allow_html: yes
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "README_figs/README-")
```

## `mamm_funrar`

This repository contains a worked through example of the use of [`funrar`R package](https://cran.r-project.org/package=funrar) in the context of biogeography using a publicly available dataset on North and Central American mammals [@lawing_community_2016; @lawing_data_2016].

The dataset is available through Dryad there: [doi:10.5061/dryad.9t0n8](https://doi.org/10.5061/dryad.9t0n8)

### Dependencies

The analysis strongly depends on several packages:

* [`remake`](https://github.com/richfitz/remake) to make the analysis portable and reproducible,
* [`dplyr`](https://github.com/hadley/dplyr) for data wrangling,
* [`rdryad`](https://github.com/ropensci/rdryad) to download the date from Dryad,
* [`rmarkdown`](https://github.com/rstudio/rmarkdown/) to generate the analysis report,
* [`vegan`](https://github.com/vegandevs/vegan) to compute null models for site-species matrix,
* [`ggplot2`](https://github.com/tidyverse/ggplot2) to plot figures in reports,
* [`devtools`](https://github.com/hadley/devtools) for session info diagnostics,
* [`funrar`](https://cran.r-project.org/package=funrar) to compute functional rarity indices.

### Session Info

```{r}
devtools::session_info()
```


### How to run

To run the analysis on your computer, install `remake`, make the working directory the project directory and run:

```
remake::make()
```

You will see the full analysis run on your computer.
With remake all the intermediate objects are viewable on a dependency graph:

```{r}
remake::diagram()
```


## References
