---
title: "How to compute functional rarity indices using `funrar`"
author: "Matthias Grenié"
date: "21 mai 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document aims to guide you through the use of `funrar` to compute functional rarity indices.
Before running the analysis be sure to install the different packages and download the needed datasets by running the following chunk:

```{r, eval = FALSE}
install.packages(c("funrar", "rdryad", "ggplot2"))

```

## Load Data

```{r}
dryad_archive = "lawning_data.zip"

# Create a temporary file
tempfile(dryad_archive)

# Download file
dryad_fetch(download_url(handle = dryad_handle), destfile = dryad_archive,
            mode = "wb")

# Extract archive
unzip(dryad_archive, exdir = substr(dryad_dir, 1, nchar(dryad_dir) - 1))

# Clean Up Temporary File
unlink(dryad_archive)
```


Trait data, site-species

## Compute distance matrix

Scale traits first

Gower's distance, Euclidean distance (continuous triaits, etc.)

`compute_dist_matrix()`

`ade4::dist.ktab()`


Need to scale

*NOTE:* 

## Compute rarity indices

### Notes on format

tidy() versus stack()

### Distinctiveness

Formula.

### Uniqueness

### Restrictedness

### Scarcity

### Other notes

Global Di computation, Local Ui computation,

Performance sparse matrices.

Other way of computing Ri

## Visualization

Indices distribution
Indices correlations
Functional Space
Maps

