---
title: "Example of Analyses with `funrar`"
author: "Matthias Grenié"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)

theme_set(theme_bw())
```

With `funrar` you can compute **Fun**ctional **Rar**ity indices. We are using a North American Mammals datasets to show the kind of questions you can come up with.

For example, you can try to map the hotspots of Regional Functional Uniqueness, i.e. the communities with highest number of functionally unique species (relatively to the whole species pool).

## Hotspots

### Functional Uniqueness

```{r}
library(dplyr)

site_rich = rowSums(subset_pres[, -c(1:3)])
names(site_rich) = subset_pres$GlobalID

site_coords = subset_pres[, 1:3]

scale_zero_one = function(x) {
    (x - min(x))/diff(range(x))
}
# Function to get average index per site
# and output a data.frame with site coords and average index
make_index_df = function(mammal_funrar, index = "Ui", zero_one = FALSE) {
    species_index = mammal_funrar[[index]][[index]]
    
    # If we want to scale index between zero and one
    if (zero_one) {
        species_index = scale_zero_one(species_index) 
    }
    
    names(species_index) = mammal_funrar[[index]]$species
    
    index_name = paste0("site_", index)
    
    site_index = data.frame(
        given_index = (as.matrix(subset_pres[, -c(1:3)]) %*%
                                      species_index) / site_rich,
        GlobalID = site_coords$GlobalID)
    
    colnames(site_index)[1] = index_name
    
    site_index = site_index %>%
        filter_(paste0("!is.na(", site_index,")")) %>%
        inner_join(site_coords, by = "GlobalID")
    
    return(site_index)
}

# Compute Functional Uniqueness per site
site_ui = make_index_df(mammal_funrar, "Ui")
site_std_ui = make_index_df(mammal_funrar, "Ui", zero_one = TRUE)
```


```{r}
ui_map = ggplot(site_ui, aes(x = Longitude, y = Latitude, z = site_Ui)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ui$Longitude),
              ylim = range(site_ui$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Uniqueness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1) +
    borders(xlim = c(-170, -23), ylim = c(9.5,83), size = 0.01,
            colour = "gray65")

```

```{r}
std_ui_map = ggplot(site_std_ui, aes(x = Longitude, y = Latitude, z = site_Ui)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ui$Longitude),
              ylim = range(site_ui$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Uniqueness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1) +
    borders(xlim = c(-170, -23), ylim = c(9.5,83), size = 0.01,
            colour = "gray65")

```

```{r, fig.height=8}
cowplot::plot_grid(ui_map, std_ui_map, align = "hv", nrow = 2)
```


### Geographical Restrictedness

```{r}
site_ri = make_index_df(mammal_funrar, "Ri")
site_std_ri = make_index_df(mammal_funrar, "Ri", zero_one = TRUE)

ri_map = ggplot(site_ri, aes(x = Longitude, y = Latitude, z = site_Ri)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ri$Longitude),
              ylim = range(site_ri$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Restrictedness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1) +
    borders(xlim = c(-170, -23), ylim = c(9.5,83), size = 0.01,
            colour = "gray65")

std_ri_map = ggplot(site_std_ri, aes(x = Longitude, y = Latitude, z = site_Ri)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ri$Longitude),
              ylim = range(site_ri$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Restrictedness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1) +
    borders(xlim = c(-170, -23), ylim = c(9.5,83), size = 0.01,
            colour = "gray65")
```

```{r, fig.height=8}
cowplot::plot_grid(ri_map, std_ri_map, align = "hv", nrow = 2)
```

### Functional Distinctiveness

```{r}
# Average Functional Distinctiveness
site_di = rowMeans(as.matrix(mammal_funrar$Di), na.rm = T) %>%
    as_data_frame() %>%
    mutate(GlobalID = rownames(.)) %>%
    rename(avg_Di = value) %>%
    inner_join(site_ri %>%
                   dplyr::select(GlobalID, Longitude, Latitude) %>%
                   mutate(GlobalID = as.character(GlobalID)),
               by = "GlobalID")

di_map = ggplot(site_di, aes(x = Longitude, y = Latitude, z = avg_Di)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_di$Longitude),
              ylim = range(site_di$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Distinctiveness per site") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1)

di_map
```



## Total distribution of indices

```{r}
species_indices = mammal_funrar$Ui %>%
    # Add Geographical Restrictedness
    inner_join(mammal_funrar$Ri, by = "species") %>%
    mutate(std_Ui = (Ui - min(Ui))/(max(Ui) - min(Ui)))

summary(species_indices)
```

### Geographical Restrictedness

```{r}
divr::make_quantile_density(species_indices$Ri) +
    labs(x = "Geographical Restrictedness",
         y = "Density",
         fill = NULL)
```


### Functional Uniqueness

```{r}
divr::make_quantile_density(species_indices$Ui) +
    labs(x = "Functional Uniqueness",
         y = "Density",
         fill = NULL) +
    geom_density(data = null_flat_ui, aes(x = Ui, y = ..density..),
                 color = "cornflowerblue", size = 0.5, show.legend = FALSE)
```


### Functional Uniqueness and Geographical Restrictedness Biplot

```{r, fig.height=6.5, fig.width=7}
biplot_ui_ri = ggplot(species_indices, aes(x = std_Ui, y = Ri)) +
    geom_point() +
    labs(x = "Functional Uniqueness",
         y = "Geographical Restrictedness",
         title = "Species Regional Indices")

biplot_ui_ri
```


```{r null_models_ui_ri, fig.height=6.5, fig.width=7}
biplot_ui_ri_null = ggplot(species_indices, aes(x = Ui, y = Ri)) +
    geom_point(data = null_flat_ui, colour = "lightblue", alpha = 0.4) +
    geom_point() +
    geom_rug() +
    labs(x = "Functional Uniqueness",
         y = "Geographical Restrictedness",
         title = "Species Regional Indices with null model (shuffled traits)")

biplot_ui_ri_null
```

### Functional Distinctiveness

#### For each site-species combination

```{r}
mammal_di = as.matrix(mammal_funrar$Di)

only_di = mammal_di[!is.na(mammal_di)]

dens_di = divr::make_quantile_density(only_di) +
    labs(x     = "Functional Distinctiveness",
         y     = "Density",
         fill  = NULL,
         title = "Density of each species-site Distinctiveness value")
dens_di
```

```{r}
null_di = null_shuffled_matrix %>%
    lapply(as.matrix) %>%
    purrr::map(~.x[!is.na(.x)]) %>%
    purrr::map(~data.frame(Di = .x) %>%
                   mutate(type = "null"))

all_di = data.frame(Di = only_di) %>%
    mutate(type = "observed") %>%
    bind_rows(null_di)


ggplot(all_di, aes(x = Di, colour = type)) +
    geom_density() +
    labs(x     = "Functional Distinctiveness",
         y     = "Density",
         title = "Site-Species Di Density with null model")
```

#### Average Distinctiveness of a species across all sites

```{r}
dens_avg_di = divr::make_quantile_density(colMeans(mammal_di, na.rm = TRUE)) +
    labs(x = "Species Average Functional Distinctiveness",
         y = "Density",
         fill = NULL)

dens_avg_di
```

```{r}
null_avg_di = null_shuffled_matrix %>%
    lapply(as.matrix) %>%
    lapply(colMeans, na.rm = TRUE) %>%
    purrr::map(~data.frame(avg_Di = .x) %>%
                   mutate(type = "null"))

avg_di = data.frame(avg_Di = colMeans(mammal_di, na.rm = TRUE),
                    type = "observed") %>%
    bind_rows(null_avg_di)

ggplot(avg_di, aes(x = avg_Di, colour = type)) +
    geom_density() +
    labs(x     = "Average Distinctiveness across all Sites",
         y     = "Density",
         title = "Average Species Di Density with null model")
```


### Functional Distinctiveness and other indices biplots

```{r}
species_all_ind = mammal_di %>%
    funrar::matrix_to_stack("Di", "Site", "species") %>%
    inner_join(species_indices, by = "species")
```

#### Functional Distinctiveness vs. Functional Uniqueness

```{r}
biplot_di_ui = ggplot(species_all_ind, aes(x = Di, y = std_Ui)) +
    geom_point(alpha = 0.3) +
    labs(x = "Functional Distinctiveness",
         y = "Functional Uniqueness (scaled)",
         title = "Species Functional Indices (local vs. regional)")

biplot_di_ui
```

```{r}
# Correlation test
cor.test(species_all_ind$Di, species_all_ind$std_Ui, method =  "spearman")
```

```{r}
# Linear model
lm_di_ui = lm(std_Ui ~ Di, data = species_all_ind)
summary(lm_di_ui)

par(mfrow = c(2, 2))

plot(lm_di_ui)
```


#### Functional Distinctiveness vs. Geographical Restrictedness

```{r}
biplot_di_ri = ggplot(species_all_ind, aes(x = Di, y = Ri)) +
    geom_point(alpha = 0.3) +
    labs(x = "Functional Distinctiveness",
         y = "Functional Restrictedness")

biplot_di_ri
```

```{r}
# Correlation test
cor.test(species_all_ind$Di, species_all_ind$Ri, method =  "spearman")
```

```{r}
# Linear model
lm_di_ri = lm(Ri ~ Di, data = species_all_ind)
summary(lm_di_ri)

par(mfrow = c(2, 2))

plot(lm_di_ri)
```

## Latitudinal Gradients

### Functional Uniqueness

```{r}
latitudinal_ui = ggplot(site_ui, aes(x = Latitude, y = site_Ui)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Uniqueness",
         title = "Latitudinal Gradient of Uniqueness (n=9497)")

latitudinal_ui
```


### Geographical Restrictedness

```{r}
latitudinal_ri = ggplot(site_ri, aes(x = Latitude, y = site_Ri)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Restrictedness",
         title = "Latitudinal Gradient of Restrictedness (n=9497)")

latitudinal_ri
```


### Functional Distinctiveness

```{r}
latitudinal_di = ggplot(site_di, aes(x = Latitude, y = avg_Di)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Distinctiveness",
         title = "Latitudinal Gradient of Distinctiveness (n=9055)")

latitudinal_di
```

