---
title: "Example of Analyses with `funrar`"
author: "Matthias Grenié"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)

# Define colors for IUCN status
status_values = c("CR" = "#E41A1C",  # Red
                  "EN" = "#FF7F00",  # Orange
                  "VU" = "#FFFF33",  # Yellow
                  "NT" = "#377EB8",  # Blue
                  "LC" = "#4DAF4A",  # Green
                  "DD" = "#984EA3",  # Purple
                  "NO" = "gray65")   # Gray

theme_set(theme_bw())
```

With `funrar` you can compute **Fun**ctional **Rar**ity indices. We are using a North American Mammals datasets to show the kind of questions you can come up with.

For example, you can try to map the hotspots of Regional Functional Uniqueness, i.e. the communities with highest number of functionally unique species (relatively to the whole species pool).

## Hotspots

### Functional Uniqueness

```{r}
library(dplyr)

# Compute Functional Uniqueness per site
site_rich = rowSums(subset_pres[, -c(1:3)])
names(site_rich) = subset_pres$GlobalID

site_coords = subset_pres[, 1:3]

make_index_df = function(mammal_funrar, index = "Ui") {
    species_index = mammal_funrar[[index]][[index]]

    names(species_index) = mammal_funrar[[index]]$species
    
    index_name = paste0("site_", index)
    
    site_index = data.frame(
        given_index = (as.matrix(subset_pres[, -c(1:3)]) %*%
                                      species_index) / site_rich,
        GlobalID = site_coords$GlobalID)
    
    colnames(site_index)[1] = index_name
    
    site_index = site_index %>%
        filter_(paste0("!is.na(", site_index,")")) %>%
        inner_join(site_coords, by = "GlobalID")
    
    return(site_index)
}

site_ui = make_index_df(mammal_funrar, "Ui")
```


```{r}
ui_map = ggplot(site_ui, aes(x = Longitude, y = Latitude, z = site_Ui)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ui$Longitude),
              ylim = range(site_ui$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Uniqueness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1) +
    borders(xlim = c(-170, -23), ylim = c(9.5,83), size = 0.01,
            colour = "gray65")

ui_map
```


### Geographical Restrictedness

```{r}
site_ri = make_index_df(mammal_funrar, "Ri")

ri_map = ggplot(site_ri, aes(x = Longitude, y = Latitude, z = site_Ri)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ri$Longitude),
              ylim = range(site_ri$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Restrictedness") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1)

ri_map
```

### Functional Distinctiveness

```{r}
# Average Functional Distinctiveness
site_di = rowMeans(as.matrix(mammal_funrar$Di), na.rm = T) %>%
    as_data_frame() %>%
    mutate(GlobalID = rownames(.)) %>%
    rename(avg_Di = value) %>%
    inner_join(site_ri %>%
                   dplyr::select(GlobalID, Longitude, Latitude) %>%
                   mutate(GlobalID = as.character(GlobalID)),
               by = "GlobalID")

di_map = ggplot(site_di, aes(x = Longitude, y = Latitude, z = avg_Di)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_di$Longitude),
              ylim = range(site_di$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Distinctiveness per site") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1)

di_map
```




## Relation between IUCN status and Functional Rarity Indices


### with Geographical Restrictedness

```{r}
species_status = mammal_funrar$Ui %>%
    # Add IUCN status
    inner_join(subset_iucn_status %>%
                   dplyr::select(TaxonName, IUCN_status) %>%
                   mutate(species = gsub(" ", "_", TaxonName)) %>%
                   dplyr::select(-TaxonName),
               by = "species") %>%
    # Add Geographical Restrictedness
    inner_join(mammal_funrar$Ri, by = "species") %>%
    mutate(IUCN_status = ifelse(is.na(IUCN_status), "NO",
                                as.character(IUCN_status))) %>%
    mutate(std_Ui = (Ui - min(Ui))/(max(Ui) - min(Ui)))

species_status$IUCN_status = factor(species_status$IUCN_status,
                                    levels = names(status_values))

summary(species_status)
```


```{r}
ggplot(species_status, aes(x = IUCN_status, y = Ri)) +
    geom_boxplot(aes(fill = IUCN_status)) +
    scale_fill_manual(values = status_values) +
    labs(x = "IUCN Status",
         y = "Geographical Restrictedness") +
    theme(legend.position = "none")
```

### with Functional Uniqueness

```{r}
ggplot(species_status, aes(x = IUCN_status, y = std_Ui)) +
    geom_boxplot(aes(fill = IUCN_status), outlier.colour = NA) +
    geom_jitter(alpha = 0.5) +
    scale_fill_manual(values = status_values) +
    labs(x = "IUCN Status",
         y = "Functional Uniqueness (scaled)") +
    theme(legend.position = "none")
```


## Total distribution of indices

### Geographical Restrictedness

```{r}
divr::make_quantile_density(species_status$Ri) +
    labs(x = "Geographical Restrictedness",
         y = "Density",
         fill = NULL)
```


### Functional Uniqueness

```{r}
divr::make_quantile_density(species_status$Ui) +
    labs(x = "Functional Uniqueness",
         y = "Density",
         fill = NULL) +
    geom_density(data = null_flat_ui, aes(x = Ui, y = ..density..),
                 color = "cornflowerblue", size = 0.5, show.legend = FALSE)
```


### Biplot

```{r, fig.height=6.5, fig.width=7}
biplot_ui_ri = ggplot(species_status, aes(x = std_Ui, y = Ri)) +
    geom_point() +
    labs(x = "Functional Uniqueness",
         y = "Geographical Restrictedness",
         title = "Species Regional Indices")

biplot_ui_ri
```


```{r null_models_ui_ri, fig.height=6.5, fig.width=7}
biplot_ui_ri_null = ggplot(species_status, aes(x = Ui, y = Ri)) +
    geom_point(data = null_flat_ui, colour = "lightblue", alpha = 0.4) +
    geom_point() +
    geom_rug() +
    labs(x = "Functional Uniqueness",
         y = "Geographical Restrictedness",
         title = "Species Regional Indices with null model (shuffled traits)")

biplot_ui_ri_null
```


#### With IUCN status

```{r, fig.height=6.5, fig.width=7}
ggplot(species_status, aes(x = std_Ui, y = Ri, color = IUCN_status)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = status_values) +
    labs(x = "Functional Uniqueness",
         y = "Geographical Restrictedness",
         title = "Species Regional Indices")
```


### Functional Distinctiveness

#### For each site-species combination

```{r}
mammal_di = as.matrix(mammal_funrar$Di)

only_di = mammal_di[!is.na(mammal_di)]

dens_di = divr::make_quantile_density(only_di) +
    labs(x     = "Functional Distinctiveness",
         y     = "Density",
         fill  = NULL,
         title = "Density of each species-site Distinctiveness value")
dens_di
```

```{r}
null_di = null_shuffled_matrix %>%
    lapply(as.matrix) %>%
    purrr::map(~.x[!is.na(.x)]) %>%
    purrr::map(~data.frame(Di = .x) %>%
                   mutate(type = "null"))

all_di = data.frame(Di = only_di) %>%
    mutate(type = "observed") %>%
    bind_rows(null_di)


ggplot(all_di, aes(x = Di, colour = type)) +
    geom_density() +
    labs(x     = "Functional Distinctiveness",
         y     = "Density",
         title = "Site-Species Di Density with null model")
```

#### Average Distinctiveness of a species across all sites

```{r}
dens_avg_di = divr::make_quantile_density(colMeans(mammal_di, na.rm = TRUE)) +
    labs(x = "Species Average Functional Distinctiveness",
         y = "Density",
         fill = NULL)

dens_avg_di
```

```{r}
null_avg_di = null_shuffled_matrix %>%
    lapply(as.matrix) %>%
    lapply(colMeans, na.rm = TRUE) %>%
    purrr::map(~data.frame(avg_Di = .x) %>%
                   mutate(type = "null"))

avg_di = data.frame(avg_Di = colMeans(mammal_di, na.rm = TRUE),
                    type = "observed") %>%
    bind_rows(null_avg_di)

ggplot(avg_di, aes(x = avg_Di, colour = type)) +
    geom_density() +
    labs(x     = "Average Distinctiveness across all Sites",
         y     = "Density",
         title = "Average Species Di Density with null model")
```



### Functional Uniqueness

```{r}
latitudinal_ui = ggplot(site_ui, aes(x = Latitude, y = site_Ui)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Uniqueness",
         title = "Latitudinal Gradient of Uniqueness (n=9497)")

latitudinal_ui
```


### Geographical Restrictedness

```{r}
latitudinal_ri = ggplot(site_ri, aes(x = Latitude, y = site_Ri)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    geom_smooth(method = "lm", color = "darkred") +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Restrictedness",
         title = "Latitudinal Gradient of Restrictedness (n=9497)")

latitudinal_ri
```


### Functional Distinctiveness

```{r}
latitudinal_di = ggplot(site_di, aes(x = Latitude, y = avg_Di)) +
    geom_point(alpha = 0.3, color = "darkblue") +
    geom_smooth() +
    labs(x     = expression('Latitude ('*degree*'N)'),
         y     = "Average Site Distinctiveness",
         title = "Latitudinal Gradient of Distinctiveness (n=9055)")

latitudinal_di
```

