---
title: "Example of Analyses with `funrar`"
author: "Matthias Grenié"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)


status_values = c("CR" = "#E41A1C",  # Red
                  "EN" = "#FF7F00",  # Orange
                  "VU" = "#FFFF33",  # Yellow
                  "NT" = "#377EB8",  # Blue
                  "LC" = "#4DAF4A",  # Green
                  "DD" = "#984EA3",  # Purple
                  "NO" = "gray65")   # Gray

theme_set(theme_bw())
                  
```

With `funrar` you can compute **Fun**ctional **Rar**ity indices. We are using a North American Mammals datasets to show the kind of questions you can come up with.

For example, you can try to map the hotspots of Regional Functional Uniqueness, i.e. the communities with highest number of functionally unique species (relatively to the whole species pool).

## Hotspots

### Functional Uniqueness

```{r}
library(assertr)
library(dplyr)

# Compute Functional Uniqueness per site
site_rich = rowSums(subset_pres[, -c(1:3)])
names(site_rich) = subset_pres$GlobalID

site_coords = subset_pres[, 1:3]

make_index_df = function(mammal_funrar, index = "Ui") {
    species_index = mammal_funrar[[index]][[index]]

    names(species_index) = mammal_funrar[[index]]$species
    
    index_name = paste0("site_", index)
    
    site_index = data.frame(
        given_index = (as.matrix(subset_pres[, -c(1:3)]) %*%
                                      species_index) / site_rich,
        GlobalID = site_coords$GlobalID)
    
    colnames(site_index)[1] = index_name
    
    site_index = site_index %>%
        verify(nrow(.) == nrow(subset_pres)) %>%
        filter_(paste0("!is.na(", site_index,")")) %>%
        inner_join(site_coords, by = "GlobalID")
    
    return(site_index)
}

site_ui = make_index_df(mammal_funrar, "Ui")
```


```{r}
ggplot(site_ui, aes(x = Longitude, y = Latitude, z = site_Ui)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ui$Longitude),
              ylim = range(site_ui$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Uniqueness")
```


### Geographical Restrictedness

```{r}
site_ri = make_index_df(mammal_funrar, "Ri")

ggplot(site_ri, aes(x = Longitude, y = Latitude, z = site_Ri)) +
    stat_summary_2d(bins = 150, color = NA) +
    coord_map("albers", at0 = 40, lat1 = 20, xlim = range(site_ri$Longitude),
              ylim = range(site_ri$Latitude)) +
    ggthemes::theme_map() +
    theme(legend.position = c(0.1, 0.2)) +
    labs(fill = "Average Restrictedness")
```


## Total distribution of indices

```{r}

```



## Relation between IUCN status and Functional Rarity Indices


### with Geographical Restrictedness

```{r}
species_status = mammal_funrar$Ui %>%
    inner_join(subset_iucn_status %>%
                   dplyr::select(TaxonName, IUCN_status) %>%
                   mutate(species = gsub(" ", "_", TaxonName)) %>%
                   dplyr::select(-TaxonName),
               by = "species") %>%
    inner_join(mammal_funrar$Ri, by = "species") %>%
    mutate(IUCN_status = ifelse(is.na(IUCN_status), "NO",
                                as.character(IUCN_status)))

species_status$IUCN_status = factor(species_status$IUCN_status,
                                    levels = names(status_values))
```


```{r}
ggplot(species_status, aes(x = IUCN_status, y = Ri)) +
    geom_boxplot(aes(fill = IUCN_status)) +
    scale_fill_manual(values = status_values)
```

### with Functional Uniqueness

```{r}
ggplot(species_status, aes(x = IUCN_status, y = Ui)) +
    geom_boxplot(aes(fill = IUCN_status)) +
    scale_fill_manual(values = status_values)
```

